Persistence in functional topology: q-tameness, local-connectivity and a correction to a theorem of Morse

### 2021-03-30 main.tex
Split sections

\documentclass{amsart}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{tikz-cd}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}

\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\dprime}{{\prime\prime}}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\interior}{int}

\setlength{\textwidth}{\paperwidth}
\addtolength{\textwidth}{-1.6in}
\setlength{\textheight}{\paperheight}
\addtolength{\textheight}{-1.8in}
\calclayout

\begin{document}	
	We use the following terminology: a \textit{neighborhood} of a point is a set containing an open neighborhood of it.

	The following is a modification of a theorem of Wilder.

	\begin{lemma} \label{l: wilder}
		Let $X$ be locally compact and Hausdorff and $Y \subseteq X$ compact. Define the following properties:
		\begin{itemize}
			\item[$(r^n)$] If $K$ and $M$ are compact subsets of $X$ with $K \cap Y \subset \interior_Y(M \cap Y)$, then $\im H_n\left(K \cap Y \to M\right)$ is finite.
			\item[$(r^n_\bullet)$] If $V$ is an open neighborhood of $y \in Y$, then there exist $U$ an open neighborhood of $y$ contained in $V$ such that $\im H_n\left(U \cap Y \to V\right)$ is finite.
		\end{itemize}
		Then, $(r^{n-1})$ together with $(r^n_\bullet)$ imply $(r^n)$.
	\end{lemma}

	\begin{proof}
		Let $M$ be fixed and let $\sigma$ be the set of compact subsets $K$ with $K \cap Y \subseteq \interior_Y(M \cap Y)$ such that $K$ has a neighborhood $U_K$ in $X$ such that $\im H_n\left(U_K \cap Y \to M\right)$ is finite.

		Recall the following fact:
		$(\ast)$ A space $S$ if locally compact and Hausdorff if and only if for every point $p$ of $S$, any neighborhood of $p$ contains a compact neighborhood of $p$.

		The collection $\sigma$ is not empty, in fact, it contains a compact neighborhood of any $y \in Y$. This follows from taking $V = X$ in $(r_\bullet^n)$ and using $(\ast)$ on the resulting open neighborhood.

		Notice that the collection $\sigma$ is closed under intersections. In fact, a stronger statement holds: if $K_1$ is compact and $K_2 \in \sigma$ then $K_1 \cap K_2 \in \sigma$. This is a consequence of the fact that the image of a map is finite can be deduced from that property holding for any map that factors it. We will next show that $\sigma$ is also closed under finite unions.

		For $i \in \{1, 2\}$ consider $K_i \in \sigma$ and let $K_i^\prime$ and $K_i^\dprime$ be compact neighborhoods of $K_i$ such that
		\begin{equation*}
		K_i \subseteq K_i^\dprime \subseteq \interior K_i^\prime \subseteq K_i^\prime \subseteq U_i,
		\end{equation*}
		where $\im H_n\left(U_i \cap Y \to M\right)$ is finite. 
		We can construct such compact neighborhoods using $(\ast)$ as follows:

		Consider for each $p \in K_i$ a compact neighborhood $K^\prime_i(p) \subseteq U_i$ and let $K_i^\dprime(p) \subseteq \interior K_i^\prime(p)$ be another compact neighborhood of $p$. The collection of open sets $\interior K_i^\dprime(p)$ covers the compact $K_i$. Let $\{p_i^{1}, \dots, p^{n}_i\} \subseteq K_i$ be such that $K_i \subseteq \bigcup_{j=1}^n \interior K^\dprime_i(p^j_i)$. Define $K_i^\dag = \bigcup_{j=1}^n K^\dag_i(p^j_i)$ for $i \in \{1,2\}$ and $\dag \in \{\prime, \dprime\}$.

		Furthermore, we have $K^\dprime_1 \cap K^\dprime_2 \cap Y \subseteq \interior_Y(K_1^\prime \cap K_2^\prime \cap Y)$. To see this consider $p \in K^\dprime_1 \cap K^\dprime_2 \cap Y$. By construction for $i \in \{1, 2\}$ there exist open sets $V_i \subseteq K_i^\prime$ such that $p \in V_i$. Letting $V = V_1 \cap V_2$ we have $p \in V \cap Y \subseteq K^\dprime_1 \cap K^\dprime_2 \cap Y$ implying $p \in \interior_Y(K^\dprime_1 \cap K^\dprime_2 \cap Y)$.

		\begin{equation*}
		\begin{tikzcd}
		& H_n(M) & \arrow[l] H_n(M) \oplus H_n(M) \\
		H_{n-1}(K_1^\prime \cap K_2^\prime \cap Y) & 
		H_{n}((K_1^\prime \cap Y) \cup (K_2^\prime \cap Y)) \arrow[l] \arrow[u] &
		H_{n}(K_1^\prime \cap Y) \oplus H_n(K_2^\prime \cap Y) \arrow[l] \arrow[u] \\
		H_{n-1}(K_1^\dprime \cap K_2^\dprime \cap Y) \arrow[u] & 
		H_{n}((K_1^\dprime \cap Y) \cup (K_2^\dprime \cap Y)) \arrow[l] \arrow[u]. &
		\end{tikzcd}
		\end{equation*}

		The upper right square has finite image because $\im H_n\left(K^\prime_i \cap Y \to M\right)$ is finite for each $i \in \{1, 2\}$. The bottom left square has a finite image using property $(r^{n-1})$. From standard homological algebra (cite Bredon 	Lemma 17.3) using that the middle row is exact, part of the Mayer-Vietoris sequence for the space $(K_1^\prime \cap Y) \cup (K_2^\prime \cap Y) \subseteq Y$, we conclude that $K = K_1 \cup K_2 \in \sigma$ with open neighborhood $U_K = \interior K^\dprime_1 \cup \interior K^\dprime_2$.

		Let $K$ be a compact set such that $\widetilde K = K \cap Y \subset \interior_Y(M \cap Y)$.
		Consider a finite subset $\{C_1, \dots, C_m\}$ of $\sigma$ such that $Y = \bigcup_{i=1}^m C_i \cap Y$. This set can be constructed using that $Y$ is compact and that any $y \in Y$ has a neighborhood in $\sigma$. Denoting $C = \bigcup_{i=1}^m C_i$ and $D = X \setminus \interior(C)$ we have $K = (K \cap C) \cup (K \cap D)$ which imply $K \in \sigma$ since $D \cap Y = \emptyset$ so $K \cap D \in \sigma$ trivially.
	\end{proof}

	We can now immediately prove the following result.

	\begin{theorem}
		Let $Y \subseteq X$ be a pair of compact Hausdorff spaces such that for any $p \in Y$ and any open neighborhood $V$ of $p$ in $X$ there exists an open neighborhood $U \subset V$ of $p$ such that $\im \big(\widetilde H_\bullet(U \cap Y) \to \widetilde H_\bullet(V)\big) = 0$. Then, $\im \big(H_d(Y) \to H_d(X)\big)$ is finitely generated for every $d$.
	\end{theorem}

	\begin{proof}
		Using the notation of Lemma \ref{l: wilder}, this would follow from $(r_n)$ applied to the case $K = Y$ and $M = X$. To show $(r_n)$ holds for every $n \geq 0$ we proceed by induction noticing that by assumption we have $(r^n_\bullet)$ for every $n$. 
	\end{proof}
\end{document}

\begin{lemma}
	Let $Y \subseteq X$ be a pair of compact Hausdorff spaces, then $(r^0)$ above holds. 
\end{lemma}

\begin{proof}
	Let $K$ and $M$ be compact subsets of $X$ with $\widetilde K = K \cap Y \subset \interior M$. For every $y \in \widetilde K$, considering the open neighborhood $\interior M$, let $U_y$ be the open set such that $\im r_{U_y \cap Y, M}$ is finite. Cover $\widetilde K$ with a finite number of such open sets $\{v_1, \dots v_n\}$. Since
	\begin{equation*}
	\begin{tikzcd}
	\bigoplus_i^n H_0(\widetilde{U_i}) \arrow[r] \arrow[d, "0"] & H_0(\widetilde{K}) \arrow[d] \arrow[r] & 0 \\
	\bigoplus_i^n H_0(M) \arrow[r] & H_0(M) &
	\end{tikzcd}
	\end{equation*}
	commutes, we have the desired conclusion.
	(Verify the universal coefficients theorem here recalling that $H_0$ is always free.)
\end{proof}

\paragraph{NOT \textit{Counterexample}} Let $Z = \left\{\left(0, 2^{-n}\right)\ |\ n \geq 0\right\} \cup \{(0, 0)\} \subset \R^2$. Let $X$ be a model of the suspension of $Z$ embedded in $\R^2$. The space $X$ is path-connected, Hausdorff, compact, and with infinite first homology. Furthermore, locally $H_1 = 0$. Taking $Y = X$ gives a counterexample.

\begin{equation*}
\begin{tikzpicture}[scale=5]
\foreach \i in {0,1,2,...,12}{
	\draw[line width=0.05mm] (-1,1) .. controls (0,1/2^\i) .. (1,1);
}
\end{tikzpicture}
\end{equation*}

A diagram to keep in mind
\begin{equation}
\begin{tikzcd}
\Z^{\infty} \arrow[r, "0"] & \Z^{\infty} \arrow[r, "1"] & \Z^{\infty} \arrow[r, "0"] & \Z^{\infty} \arrow[r, "1"] & \Z^{\infty} \\
\Z^{\infty} \arrow[r, "1"] \arrow[u, "0"] & \Z^{\infty} \arrow[r, "0"] \arrow[u, "0"] & \Z^{\infty} \arrow[r, "1"] \arrow[u, "1"] & \Z^{\infty} \arrow[r, "0"] \arrow[u, "0"] & \Z^{\infty} \arrow[u, "0"]
\end{tikzcd}
\end{equation}

Consider the following general diagram
\begin{equation}
\begin{tikzcd}
A_1 \arrow[r, "s_1"] & A_2 \arrow[r, "s_2"] & B \arrow[r, "t_1"] & C_1 \arrow[r, "t_2"] & C_2 \\
X_1 \arrow[r, "g_1"] \arrow[u, "0"] & X_2 \arrow[r, "g_2"] \arrow[u, "0"] & Y \arrow[r, "h_1"] \arrow[u, "f"] & Z_1 \arrow[r, "h_2"] \arrow[u, "0"] & Z_2 \arrow[u, "0"]
\end{tikzcd}
\end{equation}

>>>>

\immediate\write18{bibtex \jobname}
\documentclass{amsart}
\usepackage{amsmath, amsthm, amssymb}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{tikz-cd}
%\usepackage{newtxtext,newtxmath}
\usepackage[numbers,sort,compress]{natbib}
\usepackage{graphicx}

%To-do
\usepackage[marginpar]{todo}
\makeatletter
\renewcommand{\@tododisplay}[1]{%
%\marginpar{#1}%
\textsuperscript{#1}%
}
%
\renewcommand\@displaytodo[2][\todomark]{%
\@tododisplay{{\todoformat [#1~\ref{todolbl:\thetodo}]}}%
\footnotetext[\thetodo]{\todoformat #1:~#2}%
\global\@todotoks\expandafter{\the\@todotoks\todoitem{#1}{#2}}%
\@todotrue%
}%
\renewcommand\todomark{todo}
\makeatother

% theorems
\theoremstyle{plain}
\newtheorem{thm}{Theorem}[section]
\Crefname{thm}{Theorem}{Theorems}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\Crefname{prop}{Proposition}{Propositions}
\newtheorem{claim}[thm]{Claim}
\newtheorem{conj}[thm]{Conjecture}
\Crefname{conj}{Conjecture}{Conjectures}
\theoremstyle{definition}
\newtheorem{defi}[thm]{Definition}
\newtheorem{ex}[thm]{Example}
\newtheorem{rem}[thm]{Remark}

% commands
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\HS}{\mathrm{HS}}
\newcommand{\HT}{\mathrm{HT}}
\newcommand{\HLC}{\mathrm{HLC}}
\renewcommand{\H}{\mathbb{H}^d}
\newcommand{\dprime}{{\prime\prime}}
\DeclareMathOperator{\interior}{int}
\DeclareMathOperator{\im}{im}
\newcommand\CH{\check{H}}

% margins
\setlength{\textwidth}{\paperwidth}
\addtolength{\textwidth}{-1.6in}
\setlength{\textheight}{\paperheight}
\addtolength{\textheight}{-1.8in}
\calclayout

\newcommand{\anibal}[1]{\textcolor{blue}{[Anibal: #1]}}

\begin{document}

\title{Persistence in functional topology: q-tameness, local-connectivity and a correction to a theorem of Morse}
\author{}
\date{\today}

\begin{abstract}
With applications to the study of minimal surfaces in mind, Marston Morse wrote a series of papers developing versions of his famous inequalities for a broad class of real-valued functions on metric spaces. His work is remarkable in so far as it likely represents the first use of something that can be called persistent homology in the mathematical literature. From the modern point of view, it is particularly interesting that he uses Vietoris homology in order to enforce semi-continuity of persistent homology and that he states two topological conditions that supposedly ensure q-tameness. However, he is partly incorrect: We present a filtration that satisfies one of his conditions but has non-q-tame persistent homology. We also conjecture that his other condition, for which he provides no proof, is correct and discuss possible proof strategies.
\end{abstract}

\maketitle

\input{introduction}
\input{preliminaries}
\input{connectivity}
\input{counterexample}
\input{theorem}
\input{outlook}
%\input{other}

\section*{Acknowledgements}
This research has been supported by German Research Foundation (DFG) through the Collaborative Research Center SFB/TRR 109 \emph{Discretization in Geometry and Dynamics}, the Collaborative Research Center SFB/TRR 191 \emph{Symplectic Structures in Geometry, Algebra and Dynamics}, the Cluster of Excellence EXC-2181/1 \emph{STRUCTURES}, and the Research Training Group RTG 2229 \emph{Asymptotic Invariants and Limits of Groups and Spaces}.

\bibliographystyle{abbrvnaturl}
\bibliography{biblio}

\todos
\end{document}

### 2021-03-30 preliminaries.tex
Split sections

\documentclass{amsart}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{tikz-cd}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\HS}{\mathrm{HS}}
\newcommand{\dprime}{{\prime\prime}}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\interior}{int}

\setlength{\textwidth}{\paperwidth}
\addtolength{\textwidth}{-1.4in}
\setlength{\textheight}{\paperheight}
\addtolength{\textheight}{-1.8in}
\calclayout

\newcommand{\p}{\mathrm{p}}
\newcommand{\q}{\mathrm{q}}

\begin{document}

\begin{definition}
	For $n \in \Z$ a continuous map is said to be \textit{$n$-homologically small} ---denoted $\HS_n$--- if the image of the map induced by $H_n(-)$ is finitely generated, and we say it is \textit{homologically small}, denoted $\HS$, if it is $\HS_n$ for every~$n$.
\end{definition}

\begin{definition}
	Let $X$ be a space and $f \colon X \to \R$ a function. For any $t \in \R$, the \textit{$t$-sublevel set of $f$} is defined by 
	\begin{equation*}
	X_{\leq t} = f^{-1}((-\infty, t]).
	\end{equation*}
\end{definition}

\begin{definition}
	Let $M$ be a metric space. For any $p \in M$ and $\epsilon \geq 0$ denote
	\begin{equation*}
	B_\epsilon(p) = \{q \in M\ |\ d(p,q) < \epsilon\}, \qquad
	\overline B_\epsilon(p) = \{q \in M\ |\ d(p,q) \leq \epsilon\}.
	\end{equation*}
\end{definition}

\begin{definition}
	Let $M$ be a metric space and $f \colon M \to \R$ a function.
	The space $M$ is said to be \textit{strongly locally-$f$-connected} if for each $p \in M$, $e > 0$, and $n \in \N$ there exists $\delta \in (0, e)$ such that for every $c \geq 0$ the image of the map induced by $H_n(-)$ on $B_\delta(p) \cap M_{\leq f(p)+\delta+c} \to B_e(p) \cap M_{\leq f(p)+e+c}$ is $0$.
\end{definition}

\begin{theorem} \label{t:strong local connectenss implies q-tameness}
	Let $M$ be a metric space and $f \colon M \to \R$ a function. If $M$ is strongly locally-$f$-connected and all level sets are compact, then, for any $s < t$ the inclusion $M_s \to M_t$ is $\HS$.
\end{theorem}

\begin{lemma} \label{l:commutative algebra}
	Given a commutative diagram of modules over a principal ideal domain
	\begin{equation*}
	\begin{tikzcd}
	A_{1,1} \arrow[r] & A_{1,2} & \\
	A_{2,1} \arrow[r] \arrow[u] & A_{2,2} \arrow[r] \arrow[u] & A_{2,3} \\
	& A_{3,2} \arrow[r] \arrow[u] & A_{3,3} \arrow[u]
	\end{tikzcd}
	\end{equation*}
	where the middle row is exact and both $A_{2,1} \to A_{1,1}$ and $A_{3,3} \to A_{2,3}$ have finitely generated images, then $A_{3,2} \to A_{1,2}$ does as well.
\end{lemma}

\begin{lemma} \label{l:neighborhood third}
	Let $M$ be a metric space. For any compact subset $K$ and open set $U$ with $K \subseteq U$ there exists a compact set $K^\prime$ such that
	\begin{equation*}
	K \subseteq \interior(K^\prime) \subseteq K^\prime \subseteq U.
	\end{equation*}
\end{lemma}

\begin{proof}
	For any $x \in K$ choose $\epsilon(x)$ such that $B_{\epsilon(x)}(x) \subseteq U$. We have
	\begin{equation*}
	B_{\frac{\epsilon(x)}{3}}(x) \subseteq \overline B_{\frac{\epsilon(x)}{2}}(x) \subseteq B_{\epsilon(x)}(x).
	\end{equation*}
	Since $K$ is compact, there exists a finite set $\{x_\lambda \in K\}_{\Lambda}$ such that $K \subseteq \bigcup_{\Lambda} B_{\frac{\epsilon(x_\lambda)}{3}}(x_\lambda)$. Defining $K^\prime = \bigcup_{\Lambda} \overline B_{\frac{\epsilon(x_\lambda)}{2}}(x_\lambda)$ finishes the proof.
\end{proof}

\begin{lemma} \label{l:key lemma for q-tameness}
	Let $M$ and $f$ be as in Theorem \ref{t:strong local connectenss implies q-tameness}.
	Consider sets $K, L \subseteq M$ with $K$ compact and $K \subseteq \interior(L)$. For any $s < t$ there is $\delta \in (0,\, t-s)$ such that $K \cap M_{s+\delta} \to L \cap M_{t}$ is $\HS$.
\end{lemma}

\begin{proof}
	The lemma holds for $\HS$ replaced by $\HS_{(n-1)}$ for any $n \leq 0$ since $H_{n-1}(-)$ induces the zero map. We will proceed by induction on $n$ assuming the lemma for $\HS_{(n-1)}$.

	Given a compact set $L \subseteq M$ and $s < t$ let $\Sigma_{s, t}$ be the collection of all compact subsets $K$ of $\interior(L)$ for which there exists $\delta_K > 0$ and an open neighborhood $U_K$ of $K$ such that $U_K \cap M_{s+\delta_K} \to L \cap M_{t}$ is $\HS_n$.

	We start by showing that any point in $\interior(L) \cap M_s$ has a neighborhood in $\Sigma_{s, t}$.
	Let $x \in \interior(L) \cap M_{s}$ and take $e > 0$ such that $s + e < t$ and $B_e(x) \subseteq \interior(L)$.
	By the strong local-$f$-connectivity of $M$, there exists $\delta \in (0, e)$ such that for $c = s - f(x)$ the following composition is $\HS$:
	\begin{equation*}
	\overline B_{\delta/2}(x) \cap M_{s + \delta} \to
	B_\delta(x) \cap M_{s + \delta} =
	B_\delta(x) \cap M_{f(x) + c + \delta} \to
	B_e(x) \cap M_{f(x) + c + e} \to
	L \cap M_{t}.
	\end{equation*}

	We will now show that the class $\Sigma_{s,t}$ is closed under finite unions.
	For $i \in \{1, 2\}$ let $K_i$ be in $\Sigma_{s,t}$ with $\delta_i > 0$ and $K_i \subseteq U_i$ open such that $U_{i} \cap M_{s+\delta_i} \to L \cap M_{t}$ is $\HS_n$.
	We Use Lemma \ref{l:neighborhood third} to construct sets $K_i^\prime$ such that
	\begin{equation*}
	K_i \subseteq \interior(K_i^\prime) \subseteq K_i^\prime \subseteq U_i.
	\end{equation*}
	Notice that for $\delta = \min(\delta_i)$ we have $K_i^\prime \cap M_{s+\delta} \to L \cap M_t$ is $\HS_n$.
	Additionally, the induction hypothesis implies that $K_1 \cap K_2 \cap M_s \to K_1^\prime \cap K_2^\prime \cap M_{s+\delta}$ is $\HS_{(n-1)}$.
	We therefore have the following commutative diagram satisfying the assumptions of Lemma~\ref{l:commutative algebra}:
	\begin{equation*}
	\begin{tikzcd}
	H_n(L \cap M_t) \oplus H_n(L \cap M_t) \arrow[r] &
	H_n(L \cap M_t) & \\
	H_{n}(K_1^\prime \cap M_{s+\delta}) \oplus H_n(K_2^\prime \cap M_{s+\delta}) \arrow[r] \arrow[u] & 
	H_{n}((K_1^\prime \cap M_{s+\delta}) \cup (K_2^\prime \cap M_{s+\delta})) \arrow[r] \arrow[u] &
	H_{n-1}(K_1^\prime \cap K_2^\prime \cap M_{s+\delta}) \\ & 
	H_{n}((K_1 \cup K_2) \cap M_s) \arrow[r] \arrow[u] &
	H_{n-1}(K_1 \cap K_2 \cap M_s). \arrow[u]
	\end{tikzcd}
	\end{equation*}
	We conclude that $K_1 \cup K_2 \in \Sigma_{s, t}$.
	Since any compact $K \subseteq \interior(L)$ can be expressed as a finite union of sets in $\Sigma_{s,t}$ the induction step and the lemma are proven.
\end{proof}

\begin{proof}[Proof of Theorem \ref{t:strong local connectenss implies q-tameness}]
	 It follows from applying Lemma~\ref{l:key lemma for q-tameness} to $K = M_{\leq s}$ and $L = M$.
\end{proof}

\end{document}

>>>>

\section{Preliminaries}

\begin{itemize}
	\item metric space
	\item upper semi-continuous function
	\item Persistence module
	\item q-tameness
	\item Barcode
	\item bottleneck
	\item stability
\end{itemize}

\begin{defi}
	Let $M$ be a metric space. For any $p \in M$ and $\epsilon \geq 0$ denote
	\begin{equation*}
	B_\epsilon(p) = \{q \in M\ |\ d(p,q) < \epsilon\}, \qquad
	\overline B_\epsilon(p) = \{q \in M\ |\ d(p,q) \leq \epsilon\}.
	\end{equation*}
\end{defi}

\begin{defi}
	Let $X$ be a space and $f \colon X \to \R$ a function. For any $t \in \R$, the \textit{$t$-sublevel set of $f$} is defined by 
	\begin{equation*}
	X_{\leq t} = f^{-1}((-\infty, t]).
	\end{equation*}
\end{defi}
